---
title: "ESP106-Midterm2-Sravya Kamalapuram"
output: html_document
---

### Midterm2 - Sravya Kamalapuram

Loading libaries

```{r warning=FALSE, echo=FALSE, message=FALSE}
library(terra)
```

Reading the input shape files

```{r warning=FALSE, echo=FALSE, message=FALSE}
#Reading the environment data
env = rast("env.tif")

#Reading the CA county boundaries shaped file
ca_county = vect("CA_Counties_TIGER2016.shp")
```

#### Question1: Getting the range and units from env file 
```{r warning=FALSE, echo=FALSE, message=FALSE}
env

```

The ranges are given as follows: /
Elevation: -105.000 ft to 4239 ft/
TEmperature: -3.89 C to 24.4 C /
Precipitation: 46 mm to 2287mm /

Units: 
Elevation: feet/ 
Temperature: Centigrade/ 
Precipitation: Millimeters/

#### Changing the Projections

It can be seen that the projections for both the files are different.

Environmental data has a "longlat" projection. While CA County boundaries file has "Mercator" projection. 

We change the projection of CA county boundaries file to longlat. 

``` {r warning=FALSE, echo=FALSE, message=FALSE}

#Changing the projection for CA_county file 
ca_county <- terra::project(ca_county, "+proj=longlat +datum=WGS84")

paste0("CA County data")
ca_county


```

#### Cropping the Elevation data in environmental file 

To map the elevation data for California alone, we crop the elevation in the environment data to match the extent of CA county boundaries. 

``` {r warning=FALSE, echo=FALSE, message=FALSE}

#Adding a new variable by changing the extent 
env1 = crop(env, ext(ca_county))
env1 = mask(env1,ca_county)

paste ("Extent of Elevation data:")
ext(env1)

```

#### Question 2: 
#### Mapping the elevation data for California counties 

``` {r warning=FALSE, echo=FALSE, message=FALSE}
plot(env1$elevation)
plot(ca_county, add=TRUE)

```


#### Question 3: 
#### Extract and graph the elevation data for California counties
```{r warning=FALSE, echo=FALSE, message=FALSE}
#Using the extract command to extract elevation data for CA_counties
e = extract(env1$elevation, ca_county)

#Graph: Distribution of elevation data for California
density_elev = density(e$elevation, na.rm=TRUE)
p =plot(density_elev, main = "Distribution of elevation in California")

```


#### Question 4: 

Data frame with latitude and environment data for each county 

```{r warning=FALSE, echo=FALSE, message=FALSE}
#Extract the mean for each polygon of the CA-county shape file 
env_mean <- extract(env1, ca_county,mean, na.rm =TRUE)

#Calculate the centroid for each county
cent = centroids(ca_county)

#Put centroids into a dataframe, along with the latitude and longitude values
cent_df=as.data.frame(cent,xy=TRUE)

#Appending the NAME and latitude of counties to the env_mean file 
env_mean$NAME = cent$NAME
env_mean$lat = cent$INTPTLAT

#Reordering the dataframe
env_mean = env_mean[,c(1,5,2,3,4,6)]

paste0("Dataframe with Environment data and Latitudes:")
head(env_mean)

```


#### Sorting and printing counties alphabetically
```{r warning=FALSE, echo=FALSE, message=FALSE}
#Sorting the counties
name1 = env_mean[order(env_mean$NAME), ]

#Printing the first five counties 
paste0("The first 5 counties are:")
head(name1,5)
```


```{r warning=FALSE, echo=FALSE, message=FALSE}
#Printing the last five counties 
paste0("The last 5 counties are:")
tail(name1,5)
```


#### Question 5: Printing the values of hottest, wettest and highest counties 
```{r warning=FALSE, echo=FALSE, message=FALSE}
paste0("The hottest county is: ", env_mean[which.max(env_mean$temperature),]$NAME)

paste0("The wettest county is:", env_mean[which.max(env_mean$precipitation),]$NAME)

paste0("The highest county is:", env_mean[which.max(env_mean$elevation),]$NAME)

```
#### Question 6

#### Regression of temperature vs latitude + elevation 
```{r warning=FALSE, echo=FALSE, message=FALSE }

# Converting the environment shape file to a dataframe
env_df = as.data.frame(env,xy=TRUE)

#Getting a sample of 10000 cells from the data 
env_sample = env_df[sample(nrow(env_df), 10000), ]

#Regression model for the sample data, temperature vs latitude + elevation 
lm1 = lm (formula = env_sample$temperature~env_sample$elevation+env_sample$y)

#Summary of lm1
paste0("Summary of the regression model")
summary(lm1)
```

From the summary, we see that all the estimates are significant and the R-squared value is around 0.901. Which implies that the model explains close to 90% of the information in the sample data. 

#### Question 7

The coefficient of elevation in the model is -0.004410053. That is for every one foot change in the elevation, the temperature changes by -0.004410053 Centigrade, other things being constant

The coefficient of latitude is -0.829316974. That is going one degree northward would change temprature by -0.829316974. 

One degree change in the latitude, is equivalent to 111kms. That is 1km northward is 1/111th of a degree. 

```{r warning=FALSE, echo=FALSE, message=FALSE}
#Temperature change by going 1km northward 
temp_change = (-0.829316974)*1/111

#Finding the elevation for the same change in temperature 
elev_change = temp_change/-0.004410053 # in feet 
```

##### This implies that 1.6941 feet or 0.5163meters change in elevation causes equivalent temperature effect to moving 1km northward. 